---
output: github_document
title: "powdR: Full Pattern Summation of X-Ray Powder Diffraction Data"
bibliography: bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

[![Build Status](https://travis-ci.org/benmbutler/powdR.svg?branch=master)](https://travis-ci.org/benmbutler/powdR)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/powdR)](https://CRAN.R-project.org/package=powdR)
[![Total_Downloads](https://cranlogs.r-pkg.org/badges/grand-total/powdR)](https://cran.r-project.org/package=powdR)

## Overview

`powdR` is an implementation of the full pattern summation approach to quantitative mineralogy from X-ray powder diffraction data [@Chipera2002;@Chipera2013;@Eberl2003].

`powdR` has several advantages over the excel based implementations of full pattern summation such as FULLPAT [@Chipera2002] and RockJock [@Eberl2003]. First, computation is faster and, when quantifying multiple samples, can be used in combination with other packages (e.g [`foreach`](https://cran.r-project.org/web/packages/foreach/index.html)) for parralel processing. Secondly, powdR can be run via a `shiny` web application, which offers a user friendly interface for fast and iterative mineral quantification. Lastly, R represents a powerful tool for data manipulation, allowing users to creatively adapt, pre-treat and visualise their XRPD data.

## Installation

The stable version of `powdR` is on CRAN:
```{r, eval = FALSE}
install.packages("powdR")
```

Alternatively, the development version can be downloaded from Github
```{r, eval = FALSE}
#Install devtools if you don't already have it on your machine
install.packages("devtools")

devtools::install_github('benmbutler/powdR')
```

## Usage

```{r example}
library(powdR)

#Load some soils to quantify
data(soils)

#Load a powdRlib reference library of pure patterns
data(minerals)

#Quantify a sample
q <-  fps(lib = minerals,
          smpl = soils$sandstone,
          refs = minerals$phases$phase_id,
          std = "QUA.1")

#Inspect the phase concentrations (summarised by name)
q$phases_summary

#Inspect the quantification
plot(q, wavelength = "Cu")

```

Alternatively, `plot(q, wavelength = "Cu", interactive = TRUE)` provides an interactive plot for better inspection of the fit. More detailed usage instructions are provided in the package vignette.

## The powdR Shiny app

To run `powdR` via the shiny app, use `run_powdR()`. This loads the application in your default web browser. The application has six tabs:

1. **Reference Library Builder:** Allows you to create and export a `powdRlib` reference library from two .csv files: one for the XRPD measurements, and the other for the ID, name and reference intensity ratio of each pattern.
2. **Reference Library Viewer:** Facilitates quick inspection of the phases within a `powdRlib` reference library.
3. **Reference Library Editor:** Allows the user to easily subset a `powdRlib` reference library .
4. **Full Pattern Summation:** A user friendly interface for iterative full pattern summation of single samples.
5. **Automated Full Pattern Summation:** A user friendly interface for automated full pattern summation of single samples.
6. **Results viewer:** Allows for quick inspection of results derived from full pattern summation.
7. **Results editor:** Allows for results from previous saved `powdRfps` and `powdRafps` objects (.Rdata format) to be edited via addition or removal of reference patterns to the fitting process.

##References


