---
title: "powdR: Full Pattern Summation of X-ray Powder Diffraction Data"
bibliography: bibliography.bib
author: "Benjamin Butler"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
        fig_width: 7
        fig_height: 6
vignette: >
  %\VignetteIndexEntry{powdR: Full Pattern Summation of X-ray Powder Diffraction Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
code.r{ /* Code block */
    font-size: 16px;
}
p {line-height: 1.5em;}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`powdR` is an R implementation of the full pattern summation approach to quantitative mineralogy from X-ray powder diffraction (XRPD) data. Whilst this is available in Excel spreadsheets such as FULLPAT [@Chipera2002] and RockJock [@Eberl2003], implementation in R allows for faster computation than is currently available, and provides a user-friendly `shiny` application to help with the often iterative process of mineral quantification. Furthermore, the `afps` function in `powdR` is designed to automate the full pattern summation proceduce, which is particularly advantageous in high-throughput XRPD datasets.


## Full pattern summation
A powerful property of XRPD data is that it can provide quantitative estimates of phase concentrations (mineral and amorphous). Though several methods are available for quantitative mineralogy, full pattern summation (also referred to as full pattern fitting of prior measured standards) is particularly suitable for mineral mixtures containing disordered and amorphous phases. A prime example of such mixtures is soil, where clay minerals represent disordered phases, and organic matter is present as a common X-ray amorphous constituent.

The full pattern summation implemented in `powdR` is based on the principle that an observed pattern is comprised of the sum of signals from pure individual components. A key component of this approach is a library of measured or calculated patterns of pure phases. The pure patterns should be sufficient to cover all of the phases that may be encountered in your samples, and would ideally be measured on the same instrument as the sample (although this is not always possible). To quantify a given sample, phases from the library are selected, and their relative contributions to the observed signal are optimised until an appropriate fit is achieved [@Chipera2002]. The fit is usually refined using least squares optimisation of an objective parameter. The scaled intensities of the optimised patterns are then converted to weight % using reference intensity ratios, which are a measure of diffracting power relative to a standard mineral (usually corundum).

## The `powdRlib` object
A key component of functions within `powdR` is the library of pure reference patterns. These are stored within a `powdRlib` function created using the `powdRlib()` constructor function. This constructor function builds the `powdRlib` object from two components. The first argument (`xrd_table`) is a data frame of the count intensities of the reference patterns, with their 2$\theta$ axis as the first column. Each column of the reference patterns must be a unique ID. An example of this is provided in the `minerals_xrd` data:

```{r, message = FALSE, warning = FALSE}
library(powdR)

data(minerals_xrd)

head(minerals_xrd[1:8])
```
The second argument of the `powdRlib()` function is `phases_table`, which is a data frame containing 3 columns. The first is a string of unique ID's representing each reference pattern in the data provided to the `xrd_table` argument, the second column is the name of the mineral group that this reference pattern belongs to (e.g. Quartz), the third column is the reference intensity ratio (RIR) of that reference pattern (relative to a known standard, usually corundum). The RIR's are a measure of the strength of diffraction from each reference pattern, and are used to compute phase concentrations. An example of data suitable for the `phases_table` arugment is provided in the `minerals_phases` data.

```{r, message = FALSE, warning = FALSE}
data(minerals_phases)

minerals_phases[1:8,]
```

Crucially when building the `powdRlib` object, all phase ID's in the first column of the `phase_table` must match the column names of the `xrd_table` (except the name of the first column which is the 2$\theta$ scale), for example.

```{r, message = FALSE, warning = FALSE}
identical(names(minerals_xrd[-1]),
          minerals_phases$phase_id)
```

Once created, `powdRlib` objects can easily be visualised using the `plot()` function, which contains the additional arugments `wavelength` and `refs` for specifying the X-ray wavelength and the reference patterns to plot, respectively. In all cases where `plot()` is used hereafter, the addition of `interactive = TRUE` to the function call will produce an interactive html graph.

```{r, message = FALSE, warning = FALSE}
my_lib <- powdRlib(minerals_xrd, minerals_phases)

plot(my_lib, wavelength = "Cu",
     refs = c("ALB", "DOL.1",
              "QUA.1", "GOE.2"))
```

##RockJock
There are two `powdRlib` objects provided as part of the `powdR` package. The first is `minerals` (accessed via `data(minerals)`), which is a simple and low resolution library designed to facilitate faster computation of basic examples. The second is `rockjock` (accessed via `data(rockjock)`), which is a comprehensive library of 168 reference patterns covering most phases that would be encountered in geological and soil samples. The `rockjock` library in `powdR` uses that from the original RockJock program thanks to the permission of Dennis Eberl, the author of RockJock. Each of the reference patterns in `rockjock` have been scaled to a maximum intensity of 10000 counts and the RIR's re-computed accordingly.


## `fps()`
Full pattern summation in `powdR` is provided in the `fps()` function. In the following example the full pattern summation process using `fps()` will start with a reference library containing a single quartz phase, and subsequent phases will be added by examining the residuals of the fit. If running this vignette yourself, all plots can be made interactive by:

`plot(fit, wavelength = "Cu", interactive = TRUE)`

Making them interactive aids visual interpretation of the fitted patterns, which can otherwise be difficult when using static plots.

```{r, fig.show='hold', fig.cap = "**Figure 1:** The resulting fit on a sandstone soil when `fps()` is applied using a single quartz reference pattern. The top plot displays the measured and fitted patterns, and the bottom plot displays the residuals.", message = FALSE, warning = FALSE}
library(powdR)

data(minerals)
data(soils)

#Apply summation to the sandstone sample, with quartz as the only mineral
fit <- fps(lib = minerals,
           smpl = soils$sandstone,
           refs = "QUA.1",
           std = "QUA.1",
           align = 0.2)

#plot the data to interpret the fit
plot(fit, wavelength = "Cu")
```

In this case, the fitted pattern consistently overestimates the quartz peak intensities compared to the measured pattern, but underestimates the background contribution. This can occur when X-ray amorphous phases are present within the sample. To Account for this, an organic matter phase from the reference library will be added, and the fit recomputed.

```{r, fig.show='hold', fig.cap = "**Figure 2:** The resulting full pattern fit on a sandstone soil when `fps()` is applied using reference patterns from quartz and organic matter.", message = FALSE, warning = FALSE}
#Apply fps to the sandstone sample with quartz (crystalline) and organic matter (amorphous)
fit <- fps(lib = minerals,
           smpl = soils$sandstone,
           refs = c("QUA.1", "ORG"),
           std = "QUA.1",
           align = 0.2)

#plot the data to interpret the fit
plot(fit, wavelength = "Cu")
```

Now the fit is much improved, highlighting how organic matter is an important component of this pattern. Based on the current residuals, there is a notable peak at 12.3 degrees 2theta, which is often associated with kaolinite. There are also numerous peaks either side of the main quartz peak at 26.6 degrees 2theta, which are often coincident with feldspar minerals. The next fit will therefore add kaolinite, plagioclase and potassium feldspar to the process.

```{r, fig.show='hold', fig.cap = "**Figure 3:** The resulting full pattern fit on a sandstone soil when `fps` is applied using reference patterns from quartz (2 different standards), kaolinite, plagioclase, K-feldspar and organic matter.", message = FALSE, warning = FALSE}
#Apply fps to the sandstone sample, adding kaolinite, feldspar and oligoclase to the process
fit <- fps(lib = minerals,
           smpl = soils$sandstone,
           refs = c("QUA.1",
                    "QUA.2",
                    "ORG",
                    "KAO",
                    "FEL",
                    "OLI"),
           std = "QUA.1",
           align = 0.2)

#plot the data to interpret the fit
plot(fit, wavelength = "Cu")
```

When including all these phases, pretty much all peaks are accounted for. Once the user is satisfied that an appropriate fit has been achieved, the mineral concentrations can be extracted and interpreted by looking at two elements of the output from `fps()`:

- `fit$phases`
- `fit$phases_grouped`

```{r}
#returns individual contributions from each reference pattern
fit$phases

#returns summed contributions from each mineral
fit$phases_grouped
```

All of the above and subsequent examples can be further tested using the two other XRPD measurements available in the `soils` list (`soils$limestone` and `soils$granite`), each representing soils with different mineralogies.

## `fps()` with non-negative least squares

As mentioned previously, an alternative to minimisation of an objective function (the default for `fps()`) is non-negative least squares (NNLS). This is implemented in `powdR` version 0.2.0, and provides very fast computation. Whilst NNLS can struggle with X-ray amorphous phases, its speed makes it particularly useful for fast quantitative estimates of a sample using a whole reference library at once. All phases with a coefficient equal to zero are excluded.

```{r, fig.show='hold', fig.cap = "**Figure 4:** Application of `fps()`to the sandstone soil using NNLS instead of least squares optimisation.", message = FALSE, warning = FALSE}

#Apply fps to the sandstone sample using NNLS. Note that when NNLS is selected as the solver argument, there is no need to define the refs because all phases in the library are used by default
fit <- fps(lib = minerals,
           smpl = soils$sandstone,
           solver = "NNLS",
           std = "QUA.1",
           align = 0.2)

#plot the data to interpret the fit
plot(fit, wavelength = "Cu")
```

## Automated full pattern summation: `afps()`

Since NNLS is fast and can be used to exclude phases from large libraries (based on coefficients equal to zero), it has been used in combination with a method for estimating limits of detection [@Hillier2003] to create an algorithm for automated full pattern summation, `afps()`. The rationale behind this algorithm relates to the challenge of accurately quantifying high-throughput soil XRPD datasets, which show substantial variation in mineralogy. Large reference libraries are needed to handle such datasets, thus an algorithm was needed that both selected appropriate phases from a large library, and then excluded phases estimated to be below the detection limit.

```{r, fig.show='hold', fig.cap = "**Figure 5:** Application of `afps()` to the sandstone soil.", message = FALSE, warning = FALSE}

#Apply afps to the sandstone sample using afps. Note that amorphous phases need to be specified because they are treated differently to crystalline phases in this algorithm
fit <- afps(lib = minerals,
           smpl = soils$sandstone,
           std = "QUA.1",
           amorphous = "ORG",
           align = 0.2,
           lod = 0.2)

#plot the data to interpret the fit
plot(fit, wavelength = "Cu")
```

## References
