---
title: "powdR: Full Pattern Summation of X-ray Powder Diffraction Data"
bibliography: bibliography.bib
author: "Benjamin Butler"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        number_sections: true
        fig_caption: yes
        fig_width: 7
        fig_height: 6
vignette: >
  %\VignetteIndexEntry{powdR: Full Pattern Summation of X-ray Powder Diffraction Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
code.r{ /* Code block */
    font-size: 16px;
}
p {line-height: 1.5em;}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview
`powdR` is an R implementation of the full pattern summation approach to quantitative mineralogy from X-ray powder diffraction (XRPD) data. Whilst this is available in Excel spreadsheets such as FULLPAT [@Chipera2002] and RockJock [@Eberl2003], implementation in R allows for faster computation than is currently available, and provides a user-friendly `shiny` application to help with the often iterative process of mineral quantification. Furthermore, the `afps` function in `powdR` is designed to automate the full pattern summation proceduce, which is particularly advantageous in high-throughput XRPD datasets.


# Full pattern summation
A powerful property of XRPD data is that it can provide quantitative estimates of phase concentrations (mineral and amorphous). Though several methods are available for quantitative mineralogy, full pattern summation (also referred to as full pattern fitting of prior measured standards) is particularly suitable for mineral mixtures containing disordered and amorphous phases. A prime example of such mixtures is soil, where clay minerals represent disordered phases, and organic matter is present as a common X-ray amorphous constituent.

The full pattern summation implemented in `powdR` is based on the principle that an observed pattern is comprised of the sum of signals from pure individual components. A key component of this approach is a library of measured or calculated patterns of pure phases. The pure patterns should be sufficient to cover all of the phases that may be encountered in your samples, and would ideally be measured on the same instrument as the sample (although this is not always possible). To quantify a given sample, phases from the library are selected, and their relative contributions to the observed signal are optimised until an appropriate fit is achieved [@Chipera2002]. The fit is usually refined using least squares optimisation of an objective parameter. The scaled intensities of the optimised patterns are then converted to weight % using reference intensity ratios, which are a measure of diffracting power relative to a standard mineral (usually corundum).  

---

#Using `powdR`

## The `powdRlib` object
A key component of functions within `powdR` is the library of pure reference patterns. These are stored within a `powdRlib` function created using the `powdRlib()` constructor function. `powdRlib()` builds a `powdRlib` object from two components.

The first component required to build a `powdRlib` object (specified via the `xrd_table` argument of `powdRlib()`) is a data frame of the count intensities of the reference patterns, with their 2$\theta$ axis as the first column. Each column of the reference patterns must be a unique ID. An example of this is provided in the `minerals_xrd` data:

```{r, message = FALSE, warning = FALSE}
library(powdR)

data(minerals_xrd)

head(minerals_xrd[1:8])
```
The second component required to build a `powdRlib()` object (specified via the `phases_table` argument of `powdRlib()`) is a data frame containing 3 columns. The first is a string of unique ID's representing each reference pattern in the data provided to the `xrd_table` argument. The the second column is the name of the mineral group that this reference pattern belongs to (e.g. Quartz). The third column is the reference intensity ratio (RIR) of that reference pattern (relative to a known standard, usually corundum). The RIR's are a measure of the strength of diffraction from each reference pattern, and are used to compute phase concentrations. An example of data suitable for the `phases_table` argument of `powRlib()` is provided in the `minerals_phases` data.

```{r, message = FALSE, warning = FALSE}
data(minerals_phases)

minerals_phases[1:8,]
```

Crucially when building the `powdRlib` object, all phase ID's in the first column of the `phases_table` must match the column names of the `xrd_table` (except the name of the first column which is the 2$\theta$ scale), for example.

```{r, message = FALSE, warning = FALSE}
identical(names(minerals_xrd[-1]),
          minerals_phases$phase_id)
```

Once created, `powdRlib` objects can easily be visualised using `plot()`, which for `powdRlib` objects contains arguments `wavelength` and `refs` that are used to specify the X-ray wavelength and the reference patterns to plot, respectively. **In all cases where `plot()` is used hereafter, the addition of `interactive = TRUE` to the function call will produce an interactive html graph**.

```{r, message = FALSE, warning = FALSE}
my_lib <- powdRlib(minerals_xrd, minerals_phases)

plot(my_lib, wavelength = "Cu",
     refs = c("ALB", "DOL.1",
              "QUA.1", "GOE.2"))
```

## RockJock
There are two `powdRlib` objects provided as part of the `powdR` package. The first is `minerals` (accessed via `data(minerals)`), which is a simple and low resolution library designed to facilitate fast computation of basic examples. The second is `rockjock` (accessed via `data(rockjock)`), which is a comprehensive library of 168 reference patterns covering most phases that would be encountered in geological and soil samples. The `rockjock` library in `powdR` uses data from the original RockJock program [@Eberl2003] thanks to the permission of Dennis Eberl. In `rockjock`, each of the reference patterns from the original RockJock program have been scaled to a maximum intensity of 10000 counts, and the RIR's re-computed accordingly. All `rockjock` data were analysed using Cu K$\alpha$ radiation.

```{r, message = FALSE, warning = FALSE}
data(rockjock)
```

##RockJock synthetic mixtures
To accompany the `rockjock` reference library, a list of eight synthetic mixtures from the original RockJock program [@Eberl2003] are also included in `powdR` in the `rockjock_mixtures` data (accessed via `data(rockjock_mixtures)`. Their known weights (see `?rockjock_mixtures`) can be compared to full pattern summation outputs (i.e. from `fps()` and `afps()`, detailed below) to assess accuracy.

```{r, message = FALSE, warning = FALSE}
data(rockjock_mixtures)
```

##Subsetting a `powdRlib` object
Occasionally it may be useful to subset a reference library to a smaller selection. This can be achieved using `subset()`, which for `powdRlib` objects accepts three arguments; `x`, `refs` and `mode`. The `x` argument specifies the `powdRlib` object to be subset, `refs` specifies the ID's of phases to select, and `mode` specifies whether these phases are kept (`mode = "keep"`) or removed (`mode = "remove"`).

```{r, message = FALSE, warning = FALSE}
#Remove three phases from rockjock
rockjock_1 <- subset(rockjock,
                     refs = c("QUARTZ",
                              "CORUNDUM",
                              "MG_CALCITE"),
                     mode = "remove")

#Check number of phases remaining in library
nrow(rockjock_1$phases)

#Keep three phases of rockjock
rockjock_2 <- subset(rockjock,
                     refs = c("QUARTZ",
                              "CORUNDUM",
                              "MG_CALCITE"),
                     mode = "keep")

#Check number of phases remaining
nrow(rockjock_2$phases)
```

## Full pattern summation:
Full pattern summation in `powdR` is provided via `fps()`, and an automated version provided in `afps()`. Here the `rockjock` and `rockjock_mixtures` data will be used to demonstrate the use of these functions.

### Full pattern summation **with** internal standard
In some cases samples are prepared for XRPD with an internal standard of known concentration. If this is the case, then the `std_conc` argument of `fps()` and `afps()` can be used to define the concentration of the internal standard (in weight %), which is used in combination with the reference intensity ratios to compute phase concentrations. For example, all samples in the `rockjock_mixtures` data were prepared with 20 % corundum as the internal standard.

```{r, message = FALSE, warning = FALSE}
fit_1 <- fps(lib = rockjock,
             smpl = rockjock_mixtures$Mix1,
             refs = c("ORDERED_MICROCLINE",
                      "LABRADORITE",
                      "KAOLINITE_DRY_BRANCH",
                      "MONTMORILLONITE_WYO",
                      "ILLITE_1M_RM30",
                      "CORUNDUM"),
             std = "CORUNDUM",
             std_conc = 20,
             align = 0.3)
```

Notice that phase concentrations computed exclude the contribution of the internal standard...
```{r, message = FALSE, warning = FALSE}
fit_1$phases
```

...and that the phase concentrations do not sum to 100 %.
```{r, message = FALSE, warning = FALSE}
sum(fit_1$phases$phase_percent)
```

Unlike other software where only certain phases can be used as an internal standard, any phase can be defined in `powdR`. For example, the `rockjock_mixtures$Mix5` sample contains 20 % quartz, thus adding `"QUARTZ"` as the `std` argument results in this reference pattern becoming the internal standard.

```{r, message = FALSE, warning = FALSE}
fit_2 <- fps(lib = rockjock,
             smpl = rockjock_mixtures$Mix5,
             refs = c("ORDERED_MICROCLINE",
                      "LABRADORITE",
                      "KAOLINITE_DRY_BRANCH",
                      "MONTMORILLONITE_WYO",
                      "CORUNDUM",
                      "QUARTZ"),
             std = "QUARTZ",
             std_conc = 20,
             align = 0.3)

fit_2$phases

sum(fit_2$phases$phase_percent)
```
### Full pattern summation **without** internal standard
In cases where an internal standard is not added to a sample, phase quantification can be achieved by assuming that all detectable phases can be identified and that they sum to 100 weight %. By setting the `std_conc` argument of `fps()` or `afps()` to `NA`, or leaving it out of the function call, it will be assumed that the sample has been prepared without an internal standard and the phase concentrations computed accordingly.

```{r, message = FALSE, warning = FALSE}
fit_3 <- fps(lib = rockjock,
             smpl = rockjock_mixtures$Mix1,
             refs = c("ORDERED_MICROCLINE",
                      "LABRADORITE",
                      "KAOLINITE_DRY_BRANCH",
                      "MONTMORILLONITE_WYO",
                      "ILLITE_1M_RM30",
                      "CORUNDUM"),
             std_conc = NA,
             std = "CORUNDUM",
             align = 0.3)
```

In this case the phase specified in the `std` argument is only used for sample alignment, and is included in the computed phase concentrations.

```{r, message = FALSE, warning = FALSE}
fit_3$phases
```

Furthermore, the phase concentrations sum to 100 %.
```{r, message = FALSE, warning = FALSE}
sum(fit_3$phases$phase_percent)
```

### Full pattern summation with data harmonisation
It is usually recommended that the reference library used for full pattern summation is measured on the same instrument as the sample along the same 2$\theta$ range and resolution. In some cases this is not feasible, and the reference library patterns may be from a different instrument to the sample. To allow for seamless use of samples and libraries from different instruments (measured on the same wavelength), `fps()` and `afps()` contain a `harmonise` argument (default `= TRUE`). When the sample and library contain non-identical 2$\theta$ axes, `harmonise = TRUE` will convert the data onto the same scale by determining the overlapping 2$\theta$ range and interpolating to the coarsest resolution available.

```{r, message = FALSE, warning = FALSE}
#Create a sample with a shorter 2theta axis than the library
Mix1_short <- subset(rockjock_mixtures$Mix1, tth > 10 & tth < 55)

#Reduce the resolution by selecting only odd rows of the data
Mix1_short <- Mix1_short[seq(1, nrow(Mix1_short), 2),]

fit_4 <- fps(lib = rockjock,
             smpl = Mix1_short,
             refs = c("ORDERED_MICROCLINE",
                      "LABRADORITE",
                      "KAOLINITE_DRY_BRANCH",
                      "MONTMORILLONITE_WYO",
                      "ILLITE_1M_RM30",
                      "CORUNDUM"),
             std = "CORUNDUM",
             align = 0.3)

fit_4$phases
```

## Automated full pattern summation
The selection of suitable reference patterns for full pattern summation can often be challenging and time consuming. An attempt to automate this process is provided in `afps()`, which can select appropriate reference patterns from a reference library and subsequently exclude reference patterns associated with phases below an estimated limit of detection. Such an approach is considered particularly advantageous when quantifying high-throughput XRPD datasets that display considerable mineralogical variation.

Here the `rockjock` library, containing 168 reference patterns, will be used to quantify one of the samples in the `rockjock_mixtures` data.

```{r, message = FALSE, warning = FALSE}
fit_5 <- afps(lib = rockjock,
           smpl = rockjock_mixtures$Mix1,
           std = "CORUNDUM",
           align = 0.3,
           lod = 1)

fit_5$phases_grouped
```

##Plotting
Plotting results from `fps()` or `afps()` (`powdRfps` and `powdRafps` objects, respectively) is achieved using `plot()`. Static `ggplot()` plots can be created using:
```{r, message = FALSE, warning = FALSE}
plot(fit_5, wavelength = "Cu")
```

Alternatively, interactive `ggplotly()` plots can be created by adding `interactive = TRUE` to the function call, e.g. `plot(fit_5, wavelength = "Cu", interactive = TRUE)`.

##Quantifying multiple samples
The easiest way to quantify multiple samples is with `lapply()`

```{r, message = FALSE, warning = FALSE}
multi_fit <- lapply(rockjock_mixtures[1:3], fps,
                    lib = rockjock,
                    std = "CORUNDUM",
                    refs = c("ORDERED_MICROCLINE",
                             "LABRADORITE",
                             "KAOLINITE_DRY_BRANCH",
                             "MONTMORILLONITE_WYO",
                             "ILLITE_1M_RM30",
                             "CORUNDUM",
                             "QUARTZ"),
                    align = 0.3)

names(multi_fit)
```

## References
